Trygve's Problem

### 2018-03-23 generativemodelsforscrna/outline.tex
minor clarification
better plate diagram
Added sampler stuff

The Allen Institute has developed a method for clustering, one that has apparently allowed them to distinguish 116 unique cell types using single cell RNA data.  The method follows the following iterative procedure:

@@

\begin{figure}
\includegraphics[width=0.8\textwidth]{pics/plate}
\caption{Plate diagram for the model \label{fig:plate}}
\end{figure}

>>>>

The Allen Institute has developed a method for clustering, one that has apparently allowed them to distinguish 116 unique cell types using single cell RNA data.  The method follows the following recursive procedure:

@@

\begin{figure}
\includegraphics[width=0.8\textwidth]{pics/plate}
\caption{Plate diagram for the model\label{fig:plate}.  $i$ indexes layers of the hierarchy of cell types, $n$ indexes nodes in the hierarchy of cell types,  $c$ indexes cells, and $g$ indexes genes.}
\end{figure}

### 2018-03-23 trygve/outline.tex
Added sampler stuff

@@

>>>>

To this end, we use a Dikin ellipsoid sampler to approximate uniform samples from the set of valid $q$s which are consistent with what we've observed.  These are shown in Figure \ref{fig:allenavg}.  Whether the MCMC method is mixing or not is of course up for debate.  See Appendix \ref{sec:dikin} for some details.

\begin{figure}
\includegraphics[width=0.8\textwidth]{pics/allenavg}
\caption{Allen institute data: sampling information.  The first plot is the estimated $q(y|x)$, as we have already seen in other figures.  The second plot is a single sample of $q(y|x)$,  The third plot is an average value of each entry of $q$ over many samples.  The final plot indicates the standard deviation of each entry over the samples. \label{fig:allenavg}}
\end{figure}

@@

\appendix

\section{Dikin sampler}

\label{sec:dikin}

Consider a convex polytope $T=\{x:\ Ax\leq b\}$.  We have implemented a method for sampling from this polytope, based on the paper \citep{kannan2012random}.  This method makes use of the Dikin ellipsoids, $E(x)$.  For any $x$, these are defined by

\begin{itemize}
    \item Computing the distance from $x$ to each facet of the polytope, i.e. $d_i = b_i - \sum_j A_{ij} X_j$.
    \item Constructing $\tilde A$ as $\tilde A_{ij} = A_{ij} / d_i$.
    \item Define $E(x) = \{y:\ |\tilde A (X-y)|\leq 1\}$.  
\end{itemize}

We can use these ellpsoids to efficiently sample the polytope $T$.    At each step, we have some point $X\in T$, and we would to use this point to obtain a new sample $Y$, such that by iterating this process we asymptotically obtain samples which are uniform in $T$.  Here is how we use $X$ to get $Y$:

\begin{algorithm}[H]
 \KwData{A point $X\in T$}
 \KwResult{A point $Y\in T$}
 \vspace{.1in}
  Sample a proposal $\tilde Y$, uniformly from $E(X)$\;
    \eIf{$X \in E(\tilde Y)$}{
        Sample $U \sim \mathrm{Uniform}[0,1]$\;
        \eIf{$U \leq \mathrm{Vol}(E(X))/\mathrm{Vol}(E(\tilde Y)) \leq 1$}{
            Let $Y \gets \tilde Y$\;
        }{
            Let $Y \gets X$\;
        }
    }{
        Let $Y\gets X$\;
    }
 \caption{Dikin sampler step}
\end{algorithm}

It is easy to show that the stationary distribution of the Markov chain found by iterating these Dikin sampler steps is indeed uniform on $T$.  To ensure an numerically robust method in the face of high-dimensional and nearly degenerate matrices, we take the following approach to robustly sampling from the ellipsoid:

\begin{algorithm}[H]
 \KwData{An $n\times m$ matrix $\tilde A$}
 \KwResult{A point $X$ sampled uniformly from $\{x:\ |Ax|\leq 1\}$}
 \vspace{.1in}
  Sample $Z$ as an $n$-dimensional normal variables vector\;
  Let $X$ denote the solution to the least squares problem $\min_x|\tilde Ax - Z|$\;
  Normalize $X$ by $X \gets X/|\tilde AX|$\;
  Sample $U \sim \mathrm{Uniform}[0,1]$\;
  Scale $X$ by $X \gets X \times U^{1/m}$\;
 \caption{Ellipsoid sampler}
\end{algorithm}
